Quantifying the home ground advantage with `hyper2` and `hyper3` objects

### 2025-02-16 inst/home_advantage.Rmd
whitespace and label all chunks

```{r,label=usehomeaway3a}
library("hyper2",quietly=TRUE)
M <- matrix(c(NA, A+B*1i ,C+D*1i, NA),2,2,byrow=TRUE)
teams <- c("p1","p2")
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
print(M)
```

@@

```{r,label=ha3calc,cache=TRUE}
(H <- home_away3(M,lambda=2.05))
```

@@

```{r,label=findtheevaluateh3,cache=TRUE}
H
maxp(H)
```

@@

```{r,label=testequalitynow,cache=TRUE}
equalp.test(H)
```

@@

```{r,label=estimatelambda3,cache=TRUE}
f <- function(lambda){maxp(home_away3(M,lambda=lambda),give=TRUE)$value}
f(2)
f(2.1)
lam <- seq(from=1,to=10,len=17)
Supp <- sapply(lam,f)
```

```{r,label=plotsup}
plot(log(lam),Supp-max(Supp),type="b")
```

@@

```{r,label=optimlambda,cache=TRUE}
op <- optimize(f,interval=c(2,3),maximum=TRUE)
op
```

@@

```{r,ept3,cache=TRUE}
equalp.test(home_away3(M,lambda=op$maximum))
```

@@

```{r,label=loopysuppcalc,cache=TRUE}
probs <- seq(from=0.1,to=0.8,len=2)
jj <- as.matrix(expand.grid(lam,probs))
S <- rep(NA,nrow(jj))
for(i in seq_len(nrow(jj))){
    lambda <- jj[i,1]
    p1 <- jj[i,2]
    p <- c(p1,1-p1)
    names(p) <- c("p1","p2")
   S[i] <- loglik(p=p,H=home_away3(M,lambda=lambda),log=TRUE)
}
S <- matrix(S,length(lam),length(probs))
```

```{r}
contour(lam,probs,S,levels=seq(from=-20,to=-10,by=0.5),xlab="lambda",ylab="p1")
abline(h=0.5)
```

@@

```{r,label=useternary}
library(Ternary)

@@

```{r,label=genthedata}
home_games_won <- matrix(c(
    NA, 16, 12, 11,
    19, NA, 19, 16,
    17, 12, NA, 11,
    11, 12, 12, NA),
    nrow=4,ncol=4,byrow=TRUE)

@@

```{r,label=showhomeawayusingcomplex}
home_games_won + 1i*away_games_won
```

@@

```{r,label=makehyper2,cache=TRUE}
H <- home_away(home_games_won,away_games_won)
H
options("use_alabama" = FALSE) # to avoid the stupid wmmin bug
specificp.gt.test(H,"home",0)
options("use_alabama" = TRUE) # to avoid the stupid wmmin bug
```

@@

```{r,label=homegroundAgresti,cache=TRUE}
samep.test(H,teams)
```

@@

```{r,label=showvisH,cache=TRUE}
(mt <- maxp(H))
```

@@

```{r,label=showprofilelikelihood, cache=TRUE}
home_strength <- seq(from=0.15,to=0.45,length=13)
plot(home_strength,profsupp(H,"home",home_strength),type="b",pch=16,
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

@@

```{r,label=makeAgrestimatrix}
baseball_table <- matrix(c(
NA,   4+3i, 4+2i, 4+3i, 6+1i, 4+2i, 6+0i,
3+3i, NA  , 4+2i, 4+3i, 6+0i, 6+1i, 4+3i,
2+5i, 4+3i, NA  , 2+4i, 4+3i, 4+2i, 6+0i,
3+3i, 5+1i, 2+5i, NA  , 4+3i, 4+2i, 6+1i,
5+1i, 2+5i, 3+3i, 4+2i, NA  , 5+2i, 6+0i,
2+5i, 3+3i, 3+4i, 4+3i, 4+2i, NA  , 2+4i,
2+5i, 1+5i, 1+6i, 2+4i, 1+6i, 3+4i, NA),7,7)
teams <- c("Milwaukee", "Detroit", "Toronto", "New York", "Boston","Cleveland","Baltimore")
rownames(baseball_table) <- teams
colnames(baseball_table) <- teams
M <- baseball_table # saves typing
M # for convenience
```

@@

```{r,label=makesuppfrommatrix}
baseball <- home_away(M)
baseball
```

@@

```{r,label=calculatemaxpAgresti,cache=TRUE}
baseball_maxp <- maxp(baseball)
baseball_maxp
```

@@

```{r,label=showmaxpAgresti}
dotchart(baseball_maxp,pch=16)
pie(baseball_maxp)

@@

```{r,analyseAgresti,cache=TRUE}
specificp.gt.test(baseball,"home",0)
```

```{r,Agrestiprofsupp,cache=TRUE}
home_strength <- seq(from=0.005,to=0.1,length=20)
plot(home_strength,profsupp(baseball,"home",home_strength),type="b",pch=16,
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

@@

```{r,label=testhomegroundadvantage,cache=TRUE}
samep.test(baseball,teams)
```

@@

```{r}
DB <- # Davidson and Beaver
matrix(c(
    NA, 32+18i, 36+14i, 47+03i, 47+03i,
14+36i,     NA, 34+16i, 43+07i, 46+04i,
06+44i, 14+36i,     NA, 34+16i, 40+10i,
02+48i, 07+43i, 12+38i,     NA, 28+22i,
01+49i, 02+48i, 05+45i, 10+40i,     NA
),5,5,byrow=TRUE
)
jj <- c("90g","95g","100g","105g","110g")
dimnames(DB) <- list(`@home` = jj, `@away` = jj)
DB
```

@@

```{r scotmax}
(H <- home_away3(home_away_table(a,give=FALSE),RCLF3_lambda_max))  # o$maximum ~= 1.353
maxp(H,give=1)
RCLF3 <- H
```

```{r}
plot(lam,loglike-max(loglike),type='b')
abline(h=c(0,-2))
abline(v=1)
grid()
```

@@

```{r}
mH
mH["Rangers"]
mH["home"]
RCLF3_maxp
lambda <- RCLF3_lambda_max
h2_exp <- matrix(0,4,4)
rownames(h2_exp) <- teams
colnames(h2_exp) <- teams
h3_exp <- h2_exp
n_matches <- Re(aF) + Im(aF)
for(hometeam in teams){
    for(awayteam in teams){

@@

```{r, bbdlambdaD,cache=TRUE}
aT
home_draw_away3(aT,lambda=1.888,D=0.3)
f <- function(v){
    H <- home_draw_away3(aT, lambda = v[1], D=v[2])
    maxp(H,give=1)$`Log-likelihood`
}
n <- 20
lambda_try <- seq(from=1,to=3,len=n)
D_try <- seq(from=0.1,to=0.6,len=n)
V <- as.matrix(expand.grid(lambda_try,D_try))
LL <- apply(V,1,f)
```

@@

```{r}
mH
```

>>>>

```{r, label=usehomeaway3a}
library("hyper2",quietly=TRUE)
M <- matrix(c(NA, A+B*1i ,C+D*1i, NA),2,2,byrow=TRUE)
teams <- c("p1","p2")
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
print(M)
```

@@

```{r, label=ha3calc, cache=TRUE}
(H <- home_away3(M,lambda=2.05))
```

@@

```{r, label=findtheevaluateh3, cache=TRUE}
H
maxp(H)
```

@@

```{r, label=testequalitynow, cache=TRUE}
equalp.test(H)
```

@@

```{r, label=estimatelambda3, cache=TRUE}
f <- function(lambda){maxp(home_away3(M,lambda=lambda),give=TRUE)$value}
f(2)
f(2.1)
lam <- seq(from=1,to=10,len=17)
Supp <- sapply(lam,f)
```

```{r, label=plotsup}
plot(log(lam),Supp-max(Supp),type="b")
```

@@

```{r, label=optimlambda, cache=TRUE}
op <- optimize(f,interval=c(2,3),maximum=TRUE)
op
```

@@

```{r, ept3, cache=TRUE}
equalp.test(home_away3(M,lambda=op$maximum))
```

@@

```{r, label=loopysuppcalc, cache=TRUE}
probs <- seq(from=0.1,to=0.8,len=2)
jj <- as.matrix(expand.grid(lam,probs))
S <- rep(NA,nrow(jj))
for(i in seq_len(nrow(jj))){
    lambda <- jj[i,1]
    p1 <- jj[i,2]
    p <- c(p1,1-p1)
    names(p) <- c("p1","p2")
   S[i] <- loglik(p=p,H=home_away3(M,lambda=lambda),log=TRUE)
}
S <- matrix(S,length(lam),length(probs))
```

```{r contlam}
contour(lam,probs,S,levels=seq(from=-20,to=-10,by=0.5),xlab="lambda",ylab="p1")
abline(h=0.5)
```

@@

```{r, label=useternary}
library(Ternary)

@@

```{r, label=genthedata}
home_games_won <- matrix(c(
    NA, 16, 12, 11,
    19, NA, 19, 16,
    17, 12, NA, 11,
    11, 12, 12, NA),
    nrow=4,ncol=4,byrow=TRUE)

@@

```{r, label=showhomeawayusingcomplex}
home_games_won + 1i*away_games_won
```

@@

```{r, label=makehyper2, cache=TRUE}
H <- home_away(home_games_won,away_games_won)
H
options("use_alabama" = FALSE) # to avoid the stupid wmmin bug
specificp.gt.test(H,"home",0)
options("use_alabama" = TRUE) # to avoid the stupid wmmin bug
```

@@

```{r, label=homegroundAgresti, cache=TRUE}
samep.test(H,teams)
```

@@

```{r, label=showvisH, cache=TRUE}
(mt <- maxp(H))
```

@@

```{r, label=showprofilelikelihood, cache=TRUE}
home_strength <- seq(from=0.15,to=0.45,length=13)
plot(home_strength,profsupp(H,"home",home_strength),type="b",pch=16,
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

@@

```{r, label=makeAgrestimatrix}
baseball_table <- matrix(c(
NA,   4+3i, 4+2i, 4+3i, 6+1i, 4+2i, 6+0i,
3+3i, NA  , 4+2i, 4+3i, 6+0i, 6+1i, 4+3i,
2+5i, 4+3i, NA  , 2+4i, 4+3i, 4+2i, 6+0i,
3+3i, 5+1i, 2+5i, NA  , 4+3i, 4+2i, 6+1i,
5+1i, 2+5i, 3+3i, 4+2i, NA  , 5+2i, 6+0i,
2+5i, 3+3i, 3+4i, 4+3i, 4+2i, NA  , 2+4i,
2+5i, 1+5i, 1+6i, 2+4i, 1+6i, 3+4i, NA),7,7)
teams <- c("Milwaukee", "Detroit", "Toronto", "New York", "Boston","Cleveland","Baltimore")
rownames(baseball_table) <- teams
colnames(baseball_table) <- teams
M <- baseball_table # saves typing
M # for convenience
```

@@

```{r, label=makesuppfrommatrix, cache=TRUE}
baseball <- home_away(M)
baseball
```

@@

```{r, label=calculatemaxpAgresti, cache=TRUE}
baseball_maxp <- maxp(baseball)
baseball_maxp
```

@@

```{r, label=showmaxpAgresti}
dotchart(baseball_maxp,pch=16)
pie(baseball_maxp)

@@

```{r, analyseAgresti, cache=TRUE}
specificp.gt.test(baseball,"home",0)
```

```{r, Agrestiprofsupp, cache=TRUE}
home_strength <- seq(from=0.005,to=0.1,length=20)
plot(home_strength,profsupp(baseball,"home",home_strength),type="b",pch=16,
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

@@

```{r, label=testhomegroundadvantage, cache=TRUE}
samep.test(baseball,teams)
```

@@

```{r, label=defdavidsonbeaver}
DB <- # Davidson and Beaver
matrix(c(
    NA, 32+18i, 36+14i, 47+03i, 47+03i,
14+36i,     NA, 34+16i, 43+07i, 46+04i,
06+44i, 14+36i,     NA, 34+16i, 40+10i,
02+48i, 07+43i, 12+38i,     NA, 28+22i,
01+49i, 02+48i, 05+45i, 10+40i,     NA
),5,5,byrow=TRUE
)
jj <- c("90g","95g","100g","105g","110g")
dimnames(DB) <- list(`@home` = jj, `@away` = jj)
DB
```

@@

```{r scotmax, cache=TRUE}
(H <- home_away3(home_away_table(a,give=FALSE),RCLF3_lambda_max))  # o$maximum ~= 1.353
maxp(H,give=1)
RCLF3 <- H
```

```{r plotlamloglike}
plot(lam,loglike-max(loglike),type='b')
abline(h=c(0,-2))
abline(v=1)
grid()
```

@@

```{r, label=oldfirmexpectation}
mH
mH["Rangers"]
mH["home"]
RCLF3_maxp
lambda <- RCLF3_lambda_max
h2_exp <- matrix(0,4,4)
rownames(h2_exp) <- teams
colnames(h2_exp) <- teams
h3_exp <- h2_exp
n_matches <- Re(aF) + Im(aF)
for(hometeam in teams){
    for(awayteam in teams){

@@

```{r, bbdlambdaD,cache=TRUE}
aT
home_draw_away3(aT,lambda=1.888,D=0.3)
f <- function(v){
    H <- home_draw_away3(aT, lambda = v[1], D=v[2])
    maxp(H,give=1)$`Log-likelihood`
}
n <- 2
lambda_try <- seq(from=1,to=3,len=n)
D_try <- seq(from=0.1,to=0.6,len=n)
V <- as.matrix(expand.grid(lambda_try,D_try))
LL <- apply(V,1,f)
```

@@

```{r, label=showmH}
mH
```
